---
title: "Practical session 7: RISK PREDICTION"
subtitle: "Advanced Statistics for Records Research"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```
## Research question
In this session, we will explore the dataset of 2000 participants we met in the lecture, and fit
a risk prediction model for death within 5 years, based on some simple patient
characteristics.

## Objectives
By the end of this practical, you should be able to:
1. Fit a logistic model to create risk predictions.
2. Assess model discrimination by calculating the Area Under the Curve.
3. Assess model calibration by graphing observed and predicted risks.

## Dataset and analysis
For this practical we will use a (simulated) dataset called `data_predict`. This contains
data for 2,000 patients, with information on six variables.

| Variable | Description             |
|----------|-------------------------|
| id       | Unique patient ID       |
| age      | Age (years)             |
| sbp      | Systolic Blood Pressure |
| bmi      | Body Max Index $kg/m^2$ |
| sex      | Female / Male           |
| dead     | Alive / Dead            |


```{r setup}
# Install Libraries
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyr, dplyr, ggplot2, broom)

# Load data
load("../data/data_predict.rda")
```

## Data exploration
Have a look at the data. How many participants die? What proportion are female?
What ages are these participants?

```{r data}
# Base R
summary(data_predict)

# Tidyverse (More verbose but more control)
data_predict %>% 
  group_by(dead) %>% 
  tally %>% 
  mutate(percent = n/sum(n)*100)

data_predict %>% 
  group_by(sex) %>% 
  tally %>% 
  mutate(percent = n/sum(n)*100)

data_predict %>% 
  # filter out missing observation at any variable
  filter_all(all_vars(!is.na(.))) %>% 
  summarise(n = n(),
            mean = mean(age),
            sd = sd(age),
            min = min(age),
            max = max(age))
```
About 25% of the 2,000 individuals die – this is a high-risk population. There is roughly 50% males and 50% females, aged 40-80.

### Randomly split data into training and test sets
```{r partition}
set.seed(777)

n_total <- nrow(data_predict)
proportion_train <- 0.5
n_train <- floor(proportion_train * n_total)

sample_train <- sample(1:n_total, n_train) %>% sort
sample_test <- which(!(1:n_total %in% sample_train))

data_train <- data_predict[sample_train,]
data_test <- data_predict[sample_test,]
```

### Fit model in training data
```{r fit_model}
# Fit logistic regression
model <- glm(formula = dead ~ age + sex + sbp + bmi,
             family = "binomial",
             data = data_train)

# View model output summary
summary(model)

# Get odds ratio and 95% CI
cbind(OR = coef(model), confint(model)) %>% exp


```

### Predict risk based on trained model
#### On training dataset
```{r predict_in_train}
# update data_train and data_test with predicted probabilites
data_train$prob_dead <- 
  predict(model, type = "response")

# compare predicted risk 
data_train %>% 
  group_by(dead) %>% 
  summarise(n = n (),
            mean = mean(prob_dead),
            sd = sd(prob_dead),
            min = min(prob_dead),
            max = max(prob_dead)) %>% 
  pivot_longer(-dead) %>% 
  pivot_wider(id_cols = name, names_from = dead)

```

#### On test dataset
```{r predict_in_test}
data_test$prob_dead <-
  predict(model, type = "response", newdata = data_test)

data_test %>% 
  group_by(dead) %>% 
  summarise(n = n (),
            mean = mean(prob_dead),
            sd = sd(prob_dead),
            min = min(prob_dead),
            max = max(prob_dead)) %>% 
  pivot_longer(-dead) %>% 
  pivot_wider(id_cols = name, names_from = dead)
```

```{r predict_alt}
# Alternative solution (with broom::augment() )
alt_data_train <- model %>%
  # augment creates new columns with some useful information from the model
  # .fitted = predicted values
  augment(type.predict = "response")

alt_data_test <- model %>%
  augment(type.predict = "response", newdata = data_test)

```
### Validation

#### ROC
In the training dataset, the ROC is 79%.
This means that a person who did die has a 79% probability of having a higher predicted risk (of dying) than someone who did not.
This shows the model has fairly good discrimination (ability to separate those who did and did not experience the event of interest).

```{r roc}
pacman::p_load(yardstick)

# set the second level of factor variable as the event (i.e. dead)
options(yardstick.event_first = FALSE)


# combine dataset
data_grouped <- bind_rows(data_train %>% mutate(set = "Training"),
                      data_test %>% mutate(set = "Validation")) %>% 
  group_by(set)

# Calculate ROC
data_grouped %>% 
  roc_auc(truth = dead, prob_dead)

# Visualise ROC
data_roc <- data_grouped %>% 
  roc_curve(truth = dead, prob_dead)

ggplot(data_roc, aes(x = specificity, y = sensitivity, color = set)) +
  geom_abline(slope = -1, intercept = 1,
              size = 0.4, color = "grey21", linetype = "dashed") +
  geom_line() +
  theme_minimal() 
  
```

#### Hosmer-Lemeshow
The Hosmer-Lemeshow goodness of fit table for the two (S=0 and the S=1) datasets
were very similar. Both showed evidence of a well calibrated model.

```{r gof}
pacman::p_load(generalhoslem)

gof <- data_grouped %>% 
  group_map( ~ logitgof(
    obs = .$dead,
    exp = .$prob_dead,
    g = 10
  ))

# Fix name
names(gof) <- group_keys(data_grouped)[[group_vars(data_grouped)]]

# output Goodness of Fit metrics
gof

# create GoF table
gof_table <- lapply(gof, function(x){
  cbind(x$observed, x$expected) %>%
    as_tibble(rownames = "threshold") %>% 
    mutate(group = 1:nrow(.))
})

gof_table$Training
gof_table$Validation
```

Bar graphs comparing the predicted and observed risks in the S=0 and S=1 datasets
also show good calibration (in both the training and validation data).
```{r gof_bar}
# tidy up gof_table for plotting
tidy_gof <- bind_rows(gof_table, .id = "set") %>%
  dplyr::select(set, group, Obs = y1, Exp = yhat1) %>% 
  pivot_longer(
    cols = c(Obs, Exp),
    names_to = "class"
  )

# plotting
tidy_gof %>% 
  ggplot(aes(x = group, y = value, fill = class)) +
  facet_grid(cols = vars(set)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_x_continuous(breaks = 1:10) +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.title = element_blank())

```
